---
title: "Exponential growth"
output: html_notebook
author: "Clara Kuper"
---


```{r}
## Import libraries

library(ggplot2)
```


In this notebook, we define an exponential function, generate data and fit a model of the data to the function, using maximum likelihood

```{r}
###############################
##exponential growth function##
###############################

exp_growth <- function(t,a=1,b=2,ter=1, log=FALSE){
  # This function returns the exponential after time t
  # parameters: 
  # t = time how long the function was growing (unit), no default
  # a = constant, start value of the the function (units observation at t0), default: 1
  # b = basis, growth factor of the function, default: 2 (doubling)
  # ter = time constant, time required to reach the growth factor, default: 1
  
  # Example: if this is modeling a rabbit population and 
  # we observe this population over 30 days, 
  # the initial population was 20 animals, 
  # we know that every 2 months, the population is 1.5 times the time it was before, 
  # we would give it the values:
  # t = 30 (days)
  # a = 20 (individuals)
  # b = 1.5 (factor)
  # ter = 60 (days)
  
  if (!log){
    grown_population <- a*b^(t/ter)  
  }
  else if (log){
    # log formula
    grown_population <- log(a)+(t/ter)*log(b)
  }  
  
  
  return(grown_population)
}


# test:

exp_growth(10,log=FALSE)

```

Generating a data set

```{r}
# Define parameters: how long do we observe, what growth rate to we assume etc.

make_exp_data <- function(n,t,params = c(), noise = 0.2, log=FALSE){
  # this function returns a data set with values from time 1 to time t
  # we assume that n observations are made at each t 
  # params should be a named list with names "a","b" and "ter" to feed the model
  # default for params is empty, i.e. default values for the exponential function are used
  # noise adds a random component to each data point, the size of the noise depends of the value of the data point (the larger the number, the larger the noise)
  
  # This function returns a table with n*t entries and two columns - time and population  size 
  
  # generate an empty table
  population_table <- matrix(0,nrow = n*t, ncol = 2)
  colnames(population_table) <- c("time","population")
  
  # repreat observed time points
  population_table[,1] <- rep((1:t),n)
  # fill in exponential growth
  population_table[,2] <- mapply(exp_growth,population_table[,1],MoreArgs=as.list(params,log))
  # assign errors
  error_terms          <- mapply(rnorm,1,0,(noise* population_table[,2]))
  # final values 
  population_table[,2] <- population_table[,2]+error_terms
  
  # create and return data frame
  population_table <- as.data.frame(population_table)
  
  return(population_table)
}


# define parameters
params <- c(a=100, b=2,ter=2)
# define number of observations
n      <- 50
# define number of observed time points
t      <- 10

population_table <- make_exp_data(n,t,params)

# plot data
data_plot <- ggplot(data=population_table,aes(time,population))+
  geom_point()

data_plot

```

This is our data set, next, we build a framework for ml estimation

```{r}
rmsd_exp <- function(a,b,ter,data,t,n){
  # return sum of rmsd for data set and predictions
  
  pred_params <- c(a=a, b=b, ter = ter)
  preds <- make_exp_data(n=n,t=t,pred_params,noise = 0,log = FALSE)
  
  sorted_data <- log(data$population[order(data$time)])
  sorted_preds <- log(preds$population[order(preds$time)])
  
  return(sqrt(sum((sorted_preds-sorted_data)^2)/length(sorted_preds)))
}

rmsd_exponential <- function(data,t,n,startparams = NA){
  
  # this functions writes a table for likelihood estimates
  # startparams should be given in a list of sequences
  
  if(is.na(startparams)){
    # define sequences of start parameters when they are not given by the user
    a  <- 1:100
    b  <- 1:10
    ter<- 1:10
  } else {
    a <- startparams$a
    b <- startparams$b
    ter <- startparams$ter
  }
  
  #create a large matrix
  optim_matrix <- matrix(nrow = length(a)*length(b)*length(ter),ncol = 4)
  
  # write all parameters
  optim_matrix[,1] <- mapply(rep,a, length(ter)*length(b))
  optim_matrix[,2] <- mapply(rep,b, length(ter))
  optim_matrix[,3] <- ter
  
  # estimate likelihood
  optim_matrix[,4] <- mapply(rmsd_exp,optim_matrix[,1],optim_matrix[,2],optim_matrix[,3],MoreArgs = list(data,t,n))
  
  
  # return best fit
  best_fit <- which(optim_matrix[,4]==min(optim_matrix[,4]))
  
  if (length(best_fit)>1){
    print("WARNING, more than one parameter pair maximised likelihood. Returning only one")
    best_fit <- best_fit[1]
  }
  
  colnames(optim_matrix) <- c("a","b","ter","rmsd")
  optim_matrix <- as.data.frame(optim_matrix)
  
  fitted_params <- c(a=optim_matrix$a[best_fit],b=optim_matrix$b[best_fit], ter =optim_matrix$ter[best_fit])
  print(fitted_params)
  return(optim_matrix)
}

fitted_table <- rmsd_exponential(population_table,t,50)
best_fit     <- which(fitted_table$rmsd == min(fitted_table$rmsd))

fitted_params <- c(a=fitted_table$a[best_fit[1]] ,b=fitted_table$b[best_fit[1]] ,ter=fitted_table$ter[best_fit[1]])

```

```{r}
# plot data and estimation

preds <- make_exp_data(n,t,params = fitted_params,noise=0)


plot <- ggplot(data=population_table, aes(time,population))+
  geom_point()+
  geom_line(data=preds,color = "steelblue")

plot
```


```{r}
# Load data from corona cases in Germany

covid_data <- read.csv("CovidData.csv", sep = "," )

covid_data <- covid_data[,c(4,2)]
colnames(covid_data)<- c("time","population")

starts <- c(a=list(1:2),b=list(1:5),ter=list(c(3.1,3.2,3.3,3.4,3.5,3.6,3.7,3.8,3.9)))

# fit an exponential model to the reported cases

covid_fit <- rmsd_exponential(covid_data,length(covid_data$time),1,starts)
best_fit     <- which(covid_fit$rmsd == min(covid_fit$rmsd))

covid_params <- c(a=covid_fit$a[best_fit[1]] ,b=covid_fit$b[best_fit[1]] ,ter=covid_fit$ter[best_fit[1]])


# plot data and estimation

preds <- make_exp_data(1,length(covid_data$time),params = covid_params,noise=0)


plot <- ggplot(data=covid_data, aes(time,population))+
  geom_point()+
  geom_line(data=preds,color = "steelblue")

plot

```

The results from this fit suggest that the spread of covid19 in Germany is best described with an exponential function that doubled every 3.9 days.

However, the fit is not very accurate, especially in the beginning (the case numbers seem to increase faster than the model) and towards the end (the case numbers increase slower than the model suggests.)

In the next step, I will try to implement the effect of social distancing measures.

```{r}
# Effects of contaning the virus
# For simplicity - we consider all measures together and simulate an effect that leads to longer duplication circles.




```



Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.
