---
title: "Exponential growth"
output: html_notebook
author: "Clara Kuper"
---


```{r}
## Import libraries

library(ggplot2)
```


In this notebook, we define an exponential function, generate data and fit a model of the data to the function, using maximum likelihood

```{r}
###############################
##exponential growth function##
###############################

exp_growth <- function(t,a=1,b=2,ter=1){
  # This function returns the exponential after time t
  # parameters: 
  # t = time how long the function was growing (unit), no default
  # a = constant, start value of the the function (units observation at t0), default: 1
  # b = basis, growth factor of the function, default: 2 (doubling)
  # ter = time constant, time required to reach the growth factor, default: 1
  
  # Example: if this is modeling a rabbit population and 
  # we observe this population over 30 days, 
  # the initial population was 20 animals, 
  # we know that every 2 months, the population is 1.5 times the time it was before, 
  # we would give it the values:
  # t = 30 (days)
  # a = 20 (individuals)
  # b = 1.5 (factor)
  # ter = 60 (days)
  
  grown_population <- a*b^(t/ter)
  
  return(grown_population)
}


# test:

exp_growth(10)

```

Generating a data set

```{r}
# Define parameters: how long do we observe, what growth rate to we assume etc.

make_exp_data <- function(n,t,params = c(), noise = 0.2){
  # this function returns a data set with values from time 1 to time t
  # we assume that n observations are made at each t 
  # params should be a named list with names "a","b" and "ter" to feed the model
  # default for params is empty, i.e. default values for the exponential function are used
  # noise adds a random component to each data point, the size of the noise depends of the value of the data point (the larger the number, the larger the noise)
  
  # This function returns a table with n*t entries and two columns - time and population  size 
  
  # generate an empty table
  population_table <- matrix(0,nrow = n*t, ncol = 2)
  colnames(population_table) <- c("time","population")
  
  # repreat observed time points
  population_table[,1] <- rep((1:t),n)
  # fill in exponential growth
  population_table[,2] <- mapply(exp_growth,population_table[,1],MoreArgs=as.list(params))
  # assign errors
  error_terms          <- mapply(rnorm,1,0,(noise* population_table[,2]))
  # final values 
  population_table[,2] <- population_table[,2]+error_terms
  
  # create and return data frame
  population_table <- as.data.frame(population_table)
  
  return(population_table)
}


# define parameters
params <- c(b=2,ter=2)
# define number of observations
n      <- 100
# define number of observed time points
t      <- 100

population_table <- make_exp_data(n,t,params)

# plot data
data_plot <- ggplot(data=population_table,aes(time,population))+
  geom_point()

data_plot

```

This is our data set, next, we build a framework for ml estimation

```{r}
LL_exp <- function(data,time,b,ter){
  # return sum of log likelihood density for an exponential distribution
  # data: vector of population size
  # 
  # I don't know what to enter at the second position of the function dexp .... ?dexp says it should be a "vector of rates" .... I'll just keep it as it is, but I will need help here
  
  return (sum(log(dexp(data,b/ter,log=FALSE))))
}

ml_exponential <- function(data,time,t,startparams = NA){
  
  # this functions writes a table for likelihood estimates
  # startparams should be given in a list of sequences
  
  if(is.na(startparams)){
    # define sequences of start parameters when they are not given by the user
    b  <- 1:100
    ter<- 1:100
  } else {
    b <- startparams["b"]
    ter <- startparams["ter"]
  }
  
  #create a large matrix
  optim_matrix <- matrix(nrow = length(b)*length(ter),ncol = 3)
  
  # write all parameters
  optim_matrix[,1] <- mapply(rep,b, length(ter))
  optim_matrix[,2] <- ter
  
  # estimate likelihood
  optim_matrix[,3] <- mapply(LL_exp,data,time,optim_matrix[,1],optim_matrix[,2])
  
  
  # return best fit
  best_fit <- which(optim_matrix[,3]==max(optim_matrix[,3]))
  
  if (length(best_fit)>1){
    print("WARNING, more than one parameter pair maximised likelihood. Returning only one")
    best_fit <- best_fit[1]
  }
  
  colnames(optim_matrix) <- c("b","ter","ll")
  optim_matrix <- as.data.frame(optim_matrix)
  
  return(optim_matrix)
}

fit_table <- ml_exponential(population_table$population,population_table$time,t)

fitted_params <- c(b=fit_table$b[which(fit_table$ll==max(fit_table$ll))], ter =fit_table$ter[which(fit_table$ll==max(fit_table$ll))])

print(params)
print(fitted_params)

```

```{r}
# plot data and estimation

preds <- matrix(1:100,nrow = 100, ncol=2)
preds[,2]<- mapply(exp_growth,preds[,2],MoreArgs=as.list(c(fitted_params["b"],fitted_params["ter"])))
colnames(preds)<- c("time","population")

preds <- as.data.frame(preds)


plot <- ggplot(data=population_table, aes(time,log(population)))+
  geom_point()+
  geom_line(data=preds,color = "steelblue")

plot
```


Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.
